Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2025-07-09 _manuscript/index.qmd
some revision

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
library(patchwork)
library(readr)
library(kableExtra)
library(scales)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g. @gu2024data]. However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts. Ideally, there could be a way to explicitly communciate these throught processes in the data analysis to others. Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself. For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand. Externalizing these thought processes is a practice that has potential to improve the trustworthiness of the analyses [@yu2024veridical] in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

In this paper, we conceptualize these inexplicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis. We then introduce a procedure to identify the subset of checks that best predict the occurrence of unexpected outcomes. The procedure, based on simulations of the original data, compares the accuracy and independence of different combinations of analysis validation checks. Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while independence is measures the shared information among checks. The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.

@@

The concept of data quality has been studied in the literature, with earlier work focusing on defining frameworks -- such as dimensions, attributes, and measures -- to improve data quality in database and information systems [@wang1996beyond; @batini2009methodologies; @6204995; @woodall2014classification; @cai2015challenges; @8642813]. More recently, data validation has been incorporated into frameworks like Google TensorFlow [@polyzotis2019data] to ensure the quality of data for training machine learning models, as well as for supporting business decision-making [@schelter2018automating]. With the increasing availability of open data in scientific research, the users of the data are often not the original data collector and may not be aware of all the details or nuances of the data. This encourages researchers to conduct data quality checks before beginning the analytic process, helping to avoid unexpected discoveries later on. In R, packages like `skimr` [@skimr] and `dataMaid` [@dataMaid] provide basic data screening and data quality reports, while packages such as  `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuse on providing infustructures that allow users to define customized data quality checks.

@@

From the simulated data, the accuracy branch in @fig-metric-calc refers to the set of checks' ability to detect unexpected outcomes accurately. To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used. Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes. The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$. In our binary-binary classification problem, the logic regression model can be written as follows:

@@

This objective function is optimized using simulated annealing algorithm to minimize the misclassification error and the following 6 moves are permitted during the optimization to grow or purne the tree (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node. Predictions from the logic regression model is then used, along with the observed outcome, to calculate the precision and recall of the optimized Boolean combination of checks. Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based recursive partitioning methods.

@@

While checks may score high on predictive accuracy, they may be correlated and less useful to provide actions to act on the analysis. This could happen if a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause. The independence score in @fig-metric-calc is defined as 1 - total correlation per observation.

@@

$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)}$$

A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome. To standardize this measure, the total correlation *per observation* is used in calculating the independence score.

The three metrics (precision, recall, and independence) can be combiend into a single metric using the arithmetic mean, harmonic mean, or quadratic mean. The differences among these means are minimal when the three metrics are similar. However, as the differences among the metrics increases, the harmonic mean tends to produce the smallest overall score, as it penalizes low values, while the quadratic mean tends to produce the largest score by rewarding higher values more. For simple interpretation of the score, the arithmetic mean is preferred, while in applications where the difference between precision, recall, and independence need to be penalized or rewarded more, the harmonic and quadratic mean should be considered.

@@

In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts. Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine]. For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied. In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model. The dataset that we use to develop our simulations and analysis checks is from New York City, and contains daily PM10, all-cause (non-accidental) mortality, and average temperature values from 1992--2000. We then will apply the analysis checks to similar data from 39 other cities from the NMMAPS study.

@@

Logic regression model fitted to the fourteen checks and the outcome expectation (unexpected) as the response variable. The model suggests the relationship:  (mortality-PM10 correlation < -0.03) AND (mortality-temperature correlation > -0.3 OR (there exist mortality outlier AND there exist PM10 outlier)) to predict the unexpected PM10 coefficient.
:::

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#remotes::install_github("huizezhang-sherry/adtoolbox")
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
library(patchwork)
library(readr)
library(kableExtra)
library(scales)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g. @gu2024data]. However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts. Ideally, there could be a way to explicitly communicate these thought processes in the data analysis to others. Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself. For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand. Externalizing these thought processes is a practice that has potential to improve the trustworthiness of the analyses [@yu2024veridical] in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

In this paper, we conceptualize these implicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis. We then introduce a procedure to identify the subset of checks that best predict the occurrence of unexpected outcomes. The procedure, based on simulations of the original data, compares the accuracy and independence of different combinations of analysis validation checks. Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while independence measures the shared information among checks. The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.

@@

The concept of data quality has been studied in the literature, with earlier work focusing on defining frameworks -- such as dimensions, attributes, and measures -- to improve data quality in database and information systems [@wang1996beyond; @batini2009methodologies; @6204995; @woodall2014classification; @cai2015challenges; @8642813]. More recently, data validation has been incorporated into frameworks like Google TensorFlow [@polyzotis2019data] to ensure the quality of data for training machine learning models, as well as for supporting business decision-making [@schelter2018automating]. With the increasing availability of open data in scientific research, the users of the data are often not the original data collector and may not be aware of all the details or nuances of the data. This encourages researchers to conduct data quality checks before beginning the analytic process, helping to avoid unexpected discoveries later on. In R, packages like `skimr` [@skimr] and `dataMaid` [@dataMaid] provide basic data screening and data quality reports, while packages such as  `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focus on providing infrastructures that allow users to define customized data quality checks.

@@

From the simulated data, the accuracy branch in @fig-metric-calc refers to the set of checks' ability to detect unexpected outcomes accurately. To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used. Originally developed for SNP micro-array data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes. The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$. In our binary-binary classification problem, the logic regression model can be written as follows:

@@

This objective function is optimized using simulated annealing algorithm to minimize the misclassification error and the following 6 moves are permitted during the optimization to grow or prune the tree (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node. Predictions from the logic regression model is then used, along with the observed outcome, to calculate the precision and recall of the optimized Boolean combination of checks. Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based recursive partitioning methods.

@@

While checks may score high on predictive accuracy, they may be correlated and less useful to provide actions to act on the analysis. This could happen if a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause. The independence score in @fig-metric-calc is defined as 1 minus total correlation per observation.

@@

$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},$$

where $p(.)$ is the TODO. A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome. To standardize this measure, the total correlation *per observation* is used in calculating the independence score.

The three metrics (precision, recall, and independence) can be combined into a single metric using the arithmetic mean, harmonic mean, or quadratic mean. The differences among these means are minimal when the three metrics are similar. However, as the differences among the metrics increases, the harmonic mean tends to produce the smallest overall score, as it penalizes low values, while the quadratic mean tends to produce the largest score by rewarding higher values more. For simple interpretation of the score, the arithmetic mean is preferred, while in applications where the difference between precision, recall, and independence need to be penalized or rewarded more, the harmonic and quadratic mean should be considered.

@@

In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts. Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine]. In the following example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied. In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model. The dataset that we use to develop our simulations and analysis checks is from New York City, and contains daily PM10, all-cause (non-accidental) mortality, and average temperature values from 1992--2000. We then will apply the analysis checks to similar data from 39 other cities from the NMMAPS study.

@@

Logic regression model fitted to the fourteen checks and the outcome expectation (unexpected) as the response variable. The model suggests the relationship:  (mortality-PM10 correlation < $-0.03$) AND (mortality-temperature correlation > $-0.3$ OR (there exist mortality outlier AND there exist PM10 outlier)) to predict the unexpected PM10 coefficient.
:::
